Do researchers preferentially collaborate with colleagues of the same gender?

### 2018-05-27 Gender homophilly project.Rmd
Tweaks

```{r}
simulate <- function(params){

@@

  # given the gender ratio at career stages a and b, find the proportion of collaborations that are man-man, woman-woman, or mixed gender
  man.man <- function(GR_a, GR_b) (1-GR_a) * (1-GR_b)
  woman.woman <- function(GR_a, GR_b) GR_a * GR_b
  mixed <- function(GR_a, GR_b) (1-GR_a) * GR_b + (1-GR_b) * GR_a

@@

  # Find p and q, which are used to calcualte alpha

  # p is proportion of men coauthoring with men
  p <- 1 - mixed / (2 * man.man + mixed)
  # q is proportion of women coauthoring with men
  q <- mixed / (2 * woman.woman + mixed)

@@

results <- expand.grid(
  GR.stage1 = seq(0.01, 0.7, length = 201),
  GR.stage2 = seq(0.01, 0.7, length = 201),
  prop.student.prof = c(0.01, 0.25, 0.5, 0.75, .99), 
  prop.student.student = c(0.01, 0.25, 0.5, 0.75, .99)) %>% 
  filter(GR.stage1 >= GR.stage2) %>%
  simulate() %>% 
  mutate(GR.stage1 = round(GR.stage1 * 100),
         GR.stage2 = round(GR.stage2 * 100),
         prop.student.prof = prop.student.prof * 100,
         prop.student.student = prop.student.student * 100) %>%
  filter(!duplicated(paste(GR.stage1,GR.stage2,prop.student.prof,prop.student.student)))

simulation.figure <- results %>%
  ggplot(aes(GR.stage1, GR.stage2, z = alpha, fill = alpha)) + 
  facet_grid(prop.student.prof ~ prop.student.student) + 
  geom_tile() + 
  #stat_contour(binwidth = 0.1, colour = "grey20", size = 0.3) + 
  scale_fill_distiller(name = "\u03B1",
                       palette = "RdYlBu",
                       limits = c(-1,1) * max(abs(results$alpha))) +
  scale_x_continuous(expand = c(0,0), limits = c(0,70))+
  scale_y_continuous(expand = c(0,0), limits = c(0,70))+
  xlab("% women among early-career authors") + 
  ylab("% women among established authors") +
  theme_bw() + 
  theme(panel.grid.minor = element_blank(),
        strip.background = element_blank(),
        legend.position = "bottom") + 
  ggtitle("Rows show % between-stage collaborations ", 
          subtitle = "Columns show % within-stage collaborations that involve two early-career researchers")

>>>>

```{r warning=FALSE}
simulate <- function(params){

@@

  # given the gender ratio at career stages a and b, find the 
  # proportion of collaborations that are man-man, woman-woman, or mixed gender
  man.man <- function(GR_a, GR_b) (1-GR_a) * (1-GR_b)
  woman.woman <- function(GR_a, GR_b) GR_a * GR_b
  mixed <- function(GR_a, GR_b) (1-GR_a) * GR_b + (1-GR_b) * GR_a

@@

  # Find p and q, which are used to calculate alpha:
  # p is proportion of men coauthoring with men
  p <- 1 - mixed / (2 * man.man + mixed)
  # q is proportion of women coauthoring with men
  q <- mixed / (2 * woman.woman + mixed)

@@

results <- expand.grid(
  GR.stage1 = seq(0.01, 0.7, length = 201),
  GR.stage2 = seq(0.01, 0.7, length = 201),
  prop.student.prof = c(0.01, 0.25, 0.5, 0.75, .99), 
  prop.student.student = c(0.01, 0.25, 0.5, 0.75, .99)) %>% 
  filter(GR.stage1 >= GR.stage2) %>%
  simulate() %>% 
  mutate(GR.stage1 = round(GR.stage1 * 100),
         GR.stage2 = round(GR.stage2 * 100),
         prop.student.prof = prop.student.prof * 100,
         prop.student.student = prop.student.student * 100) %>%
  filter(!duplicated(paste(GR.stage1, GR.stage2, prop.student.prof, prop.student.student)))

simulation.figure <- results %>%
  ggplot(aes(GR.stage1, GR.stage2, z = round(alpha,2), fill = alpha)) + 
  facet_grid(prop.student.prof ~ prop.student.student) + 
  geom_tile() + 
  geom_abline(colour = "grey20", size = 0.3) +
  stat_contour(breaks = c(-5:-1, 1:5)/10, colour = "grey20", size = 0.3) + 
  scale_fill_distiller(name = "\u03B1",
                       palette = "RdYlBu",
                       limits = c(-1,1) * max(abs(results$alpha))) +
  scale_x_continuous(expand = c(0,0), limits = c(0,70))+
  scale_y_continuous(expand = c(0,0), limits = c(0,70))+
  xlab("% women among early-career authors") + 
  ylab("% women among established authors") +
  theme_bw() + 
  theme(panel.grid.minor = element_blank(),
        strip.background = element_blank(),
        legend.position = "bottom") + 
  ggtitle("Rows show % between-stage collaborations ", 
          subtitle = "Columns show % within-stage collaborations that involve two early-career researchers")

### 2018-05-27 manuscript/Manuscript_Claire.Rmd
Tweaks

![The Wahlund effect can make it appear as if authors prefer to publish with same-gendered colleagues, even if no such preference exists. Here, coloured circles represent male and female authors, and coauthors are linked with lines. Across the whole set of ten papers, there is an apparent excess of same-gender collaborations. Specifically, there are six same-gender papers and only four mixed-gender papers, which is fewer than the $10\times2\times0.5\times0.5 = 5$ mixed-gender papers we would expect under the null hypothesis that authors assort randomly with respect to gender. However, within each subset, there is no evidence that authors prefer to publish with same-gendered individuals. The Wahlund effect will tend to inflate the frequency of same-sex coauthors whenever the data is composed of two or more disconnected subsets of literature with different author gender ratios; these subsets could be research disciplines, older versus newer papers, or papers from authors in different countries. \label{wahlund_plot}](../figures/Figure 1 - Wahlund figure.pdf)

@@

## Theoretical expectations for $\alpha$ when the gender ratio differs between career stages
In many STEMM subjects, the gender ratio is more skewed among established researchers relative to early-career researchers [@Shaw_2012; @Holman_2018]. We hypothesised that this skew could potentially create Wahlund effects and 'reverse' Wahlund effects. For example, imagine that the majority of collaborations are between students and professors, and that the gender ratio differs between career stages: we will then see an excess of mixed-gender coauthorships (heterophily, $\alpha < 0$), even if gender has no causal effect on the selection of coauthors. Similarly, a hypothetical field in which students work only with students, and professors with professors, would have apparent gender homophily ($\alpha > 0$).

Because we can think of no tractable method of controlling for this issue using our dataset, which contains no information on career stage. Therefore, we instead decided to derive the theoretical expectations for $\alpha$ when there is a difference in gender ratio across career stages. For simplicity, our calculations assume there are only two career stages, though the conclusions are likely to generalise to a more realistic multi-tiered career ladder. Under the null model that gender has no causal effect on collaboration, we calculated $\alpha$ for various combinations of the four free parameters, i.e. the gender ratios for early- and late-career researchers, and the relative frequency of collaborations between early-early, early-late, and late-late collaborations. We then used the theoretical expectations for $\alpha$ to qualify our main conclusions (see Results). Annotated R code used to derive the theoretical expectations can be found in the Online Supplementary Material.

@@

![When there is a difference in gender ratio between early-career and established researchers, and collaboration is non-random with respect to career stage, the null expectation for $\alpha$ deviates from zero. An excess of collaborations between career stages gives the appearance of gender heterophily (lower rows, blue areas), while an excess of within-career stage collaborations produced apparent gender homophily (upper rows, red areas). However, the conditions required for strong gender homophily are quite restrictive, making it unlikely that this issue explains all of the homophily observed in Figure 2. \label{simulation}](../figures/figure5.pdf)

@@

We also found that gender homophily was marginally stronger in 2015-2016, relative to 2005-2006. Although this trend might reflect a change in the gender preferences of researchers seeking collaborators over time, there are alternative (and probably more likely) explanations for this result. For example, this trend might result from the increasing number of women working in senior positions in STEMM over the past decade [e.g. @Long_2015; @Bendels_2018; @McKenzie_2017]. As shown in \autoref{simulation}, if enough coauthorships are between junior and senior researchers, a large gender gap between career stages can give the appearance of heterophily. As this gender gap between career stages lessens, the observed values of $\alpha$ can increase.

Our study begs two questions: what causes gender homophily in science, and are our results cause for concern? These questions are closely related. For example, some of the homophily we observed might be caused by women seeking to avoid harassment or bullying from men [REF], which is clearly very concerning. Additionally, Sheltzer and Smith [-@sheltzer_2014] concluded that 'elite' male academics (i.e. recipients of major honours) have a higher proportion of male students and postdocs than non-elite male academics. This finding could contribute to the homophily we observed, and is cause for concern because the results might reflect discrimination against women by established scientists [@Moss_2012], or avoidance by women of elite research groups (e.g. due to 'imposter syndrome' or the perception that these groups are a 'boy's club'). We also found a little evidence that gender homophily is deterimental to research quality, in that high-impact journals tended to have weaker homophily. Assuming that papers published in high-impact journals are of higher average quality [which is contentious; @Garfield_2006], then this result provides non-experimental support for the hypothesis that mixed-gender teams produce better research than single-gender teams [@Campbell_2013].

>>>>

![The Wahlund effect can make it appear as if authors prefer to publish with same-gendered colleagues, even if no such preference exists. Here, coloured circles represent male and female authors, and coauthors are linked with lines. Across the whole set of ten papers, there is an apparent excess of same-gender collaborations: there are six same-gender papers and only four mixed-gender papers, which is fewer than the $10\times2\times0.5\times0.5 = 5$ mixed-gender papers we would expect under the null hypothesis that authors assort randomly with respect to gender. However, within each subset, there is no evidence that authors prefer to publish with same-gendered individuals. The Wahlund effect will tend to inflate the frequency of same-sex coauthorships whenever the data is composed of two or more disconnected subsets of literature with different author gender ratios; these subsets could be research disciplines, older versus newer papers, or papers from authors in different countries. \label{wahlund_plot}](../figures/Figure 1 - Wahlund figure.pdf)

@@

## Theoretical expectations for $\alpha$ when the gender ratio differs between career stages
In many STEMM subjects, the gender ratio is more skewed among established researchers relative to early-career researchers [@Shaw_2012; @Holman_2018]. We hypothesised that this skew could potentially create both Wahlund effects and 'reverse' Wahlund effects. For example, imagine that the majority of collaborations are between students and professors, and that the gender ratio differs between career stages: we will then see an excess of mixed-gender coauthorships (heterophily, $\alpha < 0$), even if gender has no causal effect on the selection of coauthors. Similarly, a hypothetical field in which students work only with students, and professors with professors, would have apparent gender homophily ($\alpha > 0$).

We can think of no tractable method of controlling for this issue using our dataset, which contains no information on career stage. Therefore, we instead decided to derive the theoretical expectations for $\alpha$ when there is a difference in gender ratio across career stages, in order to determine whether this effect could appreciably affect our conclusions. For simplicity, our calculations assume there are only two career stages, though the conclusions are likely to generalise to a realistic, multi-tiered career ladder. Under the null model that gender has no causal effect on collaboration, we calculated $\alpha$ for various combinations of the four free parameters, i.e. the gender ratios for early- and late-career researchers, and the relative frequency of collaborations between early-early, early-late, and late-late collaborations. We then used the theoretical expectations for $\alpha$ to qualify our main conclusions (see Results). The Online Supplementary Material gives annotated R code used to derive the theoretical expectations.

@@

![When there is a difference in gender ratio between early-career and established researchers, and collaboration is non-random with respect to career stage, the null expectation for $\alpha$ deviates from zero. An excess of collaborations between career stages gives the appearance of gender heterophily (lower rows, blue areas), while an excess of within-career stage collaborations produced apparent gender homophily (upper rows, red areas). However, the conditions required for strong gender homophily are quite restrictive, making it unlikely that this issue explains all of the homophily observed in Figure 2. Contour lines denote increments of 0.1. \label{simulation}](../figures/figure5.pdf)

@@

We also found that gender homophily was marginally stronger in 2015-2016, relative to 2005-2006. Although this trend might reflect a change in the gender preferences of researchers seeking collaborators over time, there are alternative (and probably more likely) explanations for this result. For example, this trend might result from the increasing number of women working in senior positions in STEMM over the past decade [e.g. @Long_2015; @Bendels_2018; @McKenzie_2017]. As shown in \autoref{simulation}, if enough coauthorships are between junior and senior researchers, a large gender gap between career stages can give the appearance of heterophily. As this gender gap between career stages lessens, the observed values of $\alpha$ may increase.

Our study begs two questions: what causes gender homophily in science, and are our results cause for concern? These questions are closely related. For example, some of the homophily we observed might be caused by women seeking to avoid harassment or bullying from men [@jagsi_2016], which is clearly very concerning. Additionally, Sheltzer and Smith [-@sheltzer_2014] concluded that 'elite' male academics (i.e. recipients of major honours) have a higher proportion of male students and postdocs than non-elite male academics. This finding could contribute to the homophily we observed, and is cause for concern because the results might reflect discrimination against women by established scientists [@Moss_2012], or avoidance by women of elite research groups (e.g. due to 'imposter syndrome' or the perception that these groups are hostile to women). We also found a little evidence that gender homophily is deterimental to research quality, in that high-impact journals tended to have weaker homophily. Assuming that papers published in high-impact journals are of higher average quality [which is contentious; @Garfield_2006], then this result provides non-experimental support for the hypothesis that mixed-gender teams produce better research than single-gender teams [@Campbell_2013].
